---
title: "data-wrangling-practice"
author: "Ryan Kelley"
date: "6/24/2021"
output: html_document
---

## Data Wrangling Basics - Zillow Prize dataset

For this exercise in data wrangling, I'll use the Zillow Prize dataset from Kaggle: [Zillow Prize at Kaggle](kaggle.com/c/zillow-prize-1/data). This dataset is provided by Zillow as part of a public competition intended to improve the accuracy of "Zestimates" - estimated home values which are generated using machine learning models and provided for free to consumers. 

For now, I'll just use the dataset to practice importing and data transformation, and move on later to using the freshly transformed data to create some visualizations. 

First, I need to load packages. I'll use the **tidyverse** library to help me along. The **tidyverse** library includes many packages which make data wrangling and visualization much easier.  

```{r library}
library (tidyverse)
```

Next, I need to import the raw data from the .csv file that I've downloaded from Kaggle, _properties_2016.csv_. Let's make sure we can see it in our current directory. _getwd()_ will provide the current working directory ("wd"), and then I can use that output to pipe to _dir()_ to see what files are in it.

```{r dir}
getwd() %>% dir()
```

There it is! Now I need to pull the raw .csv data into R. I won't assign the input to anything yet - at first I just want to look at what happens when I run the .csv file through _read_csv()_. Sometimes R makes weird assumptions about what data type, etc. the columns should be, so I want to make sure it's not jumping to any odd conclusions.

``` {r read_csv attempt 1}
read_csv ("./properties_2016.csv") 
```

I immediately get the warning "12808 parsing failures", with the first 5 rows of failures listed. It looks like the **architecturalstyletypeid** and **typeconstructiontypeid** columns have some issues, in particular: there are some values of *6* and *7*, but R assumed that these columns should be logical type, which only takes 1/0/T/F/TRUE/FALSE as input. Since the columns reference a "type", I assume that the numbers used here are categories. I'll try forcing R to pull these columns in as datatype *factor*, instead, and see if that resolves the issue. (More on this in the [tidyverse docs](https://readr.tidyverse.org/articles/readr.html#column-specification-1).)

```{r read_csv parse attempt}
# using the .default flag would usually assign a specific data type to ALL columns unless otherwise specified; using "?" will preserve R's default parsing/guessing except where I override it
read_csv ("./properties_2016.csv", col_types = cols(.default = "?", architecturalstyletypeid = "f", typeconstructiontypeid = "f")) %>% glimpse()
```

The parsing errors have disappeared, so I'll assign the output to a new dataframe for more cleanup. From there, I'll do some preliminary cleanup and renaming of columns, before going into further tidying/normalization. The first thing I'll do is use the *na* argument in _read.csv()_ to make sure that any blank values are interpreted as NA. 

```{r read_csv to df}
# assigning to a dataframe and making sure blank values are interpreted as NA
df_2016props <- read_csv ("./properties_2016.csv", na = c("", " ", "NA"), col_types = cols(.default = "?", architecturalstyletypeid = "f", typeconstructiontypeid = "f"))
df_2016props %>% glimpse()
```
Next, I want to rename some of the columns to make them shorter and easier to read. This is where I'll pick up next time!