---
title: "data-wrangling-practice"
author: "Ryan Kelley"
date: "6/24/2021"
output: html_document
---

## Data Wrangling Basics - Zillow Prize dataset

For this exercise in data wrangling, I'll use the Zillow Prize dataset from Kaggle: [Zillow Prize at Kaggle](kaggle.com/c/zillow-prize-1/data). This dataset is provided by Zillow as part of a public competition intended to improve the accuracy of "Zestimates" - estimated home values which are generated using machine learning models and provided for free to consumers. 

For now, I'll just use the dataset to practice importing and data transformation, and move on later to using the freshly transformed data to create some visualizations. 

First, I need to load packages. I'll use the **tidyverse** library to help me along. The **tidyverse** library includes many packages which make data wrangling and visualization much easier.  

```{r library}
library (tidyverse)
```

Next, I need to import the raw data from the .csv file that I've downloaded from Kaggle, _properties_2016.csv_. Let's make sure we can see it in our current directory. _getwd()_ will provide the current working directory ("wd"), and then I can use that output to pipe to _dir()_ to see what files are in it.

```{r dir}
getwd() %>% dir()
```

There it is! Now I need to pull the raw .csv data into R. I won't assign the input to anything yet - at first I just want to look at what happens when I run the .csv file through _read_csv()_. Sometimes R makes weird assumptions about what data type, etc. the columns should be, so I want to make sure it's not jumping to any odd conclusions.

``` {r read_csv attempt 1}
read_csv ("./properties_2016.csv") 
```

I immediately get the warning "12808 parsing failures", with the first 5 rows of failures listed. It looks like the **architecturalstyletypeid** and **typeconstructiontypeid** columns have some issues, in particular: there are some values of *6* and *7*, but R assumed that these columns should be logical type, which only takes 1/0/T/F/TRUE/FALSE as input. Since the columns reference a "type", I assume that the numbers used here are categories. I'll try forcing R to pull these columns in as datatype *factor*, instead, and see if that resolves the issue. (More on this in the [tidyverse docs](https://readr.tidyverse.org/articles/readr.html#column-specification-1).)

```{r read_csv parse attempt}
# using the .default flag would usually assign a specific data type to ALL columns unless otherwise specified; using "?" will preserve R's default parsing/guessing except where I override it
read_csv ("./properties_2016.csv", col_types = cols(.default = "?", architecturalstyletypeid = "f", typeconstructiontypeid = "f")) %>% glimpse()
```

The parsing errors have disappeared, so I'll assign the output to a new dataframe for more cleanup. From there, I'll do some preliminary cleanup and renaming of columns, before going into further tidying/normalization. The first thing I'll do is use the *na* argument in _read.csv()_ to make sure that any blank values are interpreted as NA. 

```{r read_csv to df}
# assigning to a dataframe and making sure blank values are interpreted as NA
df_2016props <- read_csv ("./properties_2016.csv", na = c("", " ", "NA"), col_types = cols(.default = "?", architecturalstyletypeid = "f", typeconstructiontypeid = "f"))
df_2016props %>% glimpse()
```
Next, I want to change some of the data types for, e.g., columns which were imported as doubles which really should be factor (ones that indicate a "type", in other words, as opposed to a numeric value. I'll also do this for variables like unit, bedroom, and bathroom count, since those have more of a categorical vibe as well  (i.e. I'm more likely to be interested in the number of units/rooms as a *type* of house as opposed to treating it as a simply numeric value). 

Then, I'll rename some of the columns to make them easier to read - looking at the Data Dictionary included with this dataset on Kaggle, I think I can create some more descriptive column names than the default. 

After these cleanup steps, I can start doing some initial visualizations to get a sense of the shape of some of this data.

``` {r datatype cleanup}
# changing some of the default import data types
df_2016props <- df_2016props %>% 
  mutate (parcelid = as.factor (parcelid),
          airconditioningtypeid = as.factor (airconditioningtypeid),
          bathroomcnt = as.factor (bathroomcnt),
          bedroomcnt = as.factor (bedroomcnt),
          buildingclasstypeid = as.factor (buildingclasstypeid),
          buildingqualitytypeid = as.factor (buildingqualitytypeid),
          fullbathcnt = as.factor (fullbathcnt),
          garagecarcnt = as.factor (garagecarcnt),
          heatingorsystemtypeid = as.factor (heatingorsystemtypeid),
          pooltypeid10 = as.factor (pooltypeid10),
          pooltypeid7 = as.factor (pooltypeid7),
          propertylandusetypeid = as.factor (propertylandusetypeid),
          storytypeid = as.factor (storytypeid),
          unitcnt = as.factor (unitcnt),
          numberofstories = as.factor (numberofstories),
          taxdelinquencyflag = as.logical (taxdelinquencyflag)
          )
# drop unneeded columns
df_2016props <- df_2016props %>% select (!c(
  parcelid, 
  calculatedbathnbr, 
  decktypeid,
  threequarterbathnbr, 
  finishedfloor1squarefeet, 
  finishedsquarefeet13, 
  finishedsquarefeet50, 
  fips, 
  fullbathcnt, 
  garagetotalsqft, 
  parcelid, poolsizesum, 
  propertycountylandusecode, 
  propertyzoningdesc, 
  rawcensustractandblock, 
  censustractandblock, 
  typeconstructiontypeid, 
  structuretaxvaluedollarcnt))

# rename remaining cols per data dictionary descriptions
df_2016props <- 
  df_2016props %>% rename (
    AC_type = airconditioningtypeid,
    architecture_type = architecturalstyletypeid,
    basement_area = basementsqft,
    bathrooms = bathroomcnt,
    bedrooms = bedroomcnt,
    framing_type = buildingclasstypeid,
    building_condition = buildingqualitytypeid,
    total_finished_area = calculatedfinishedsquarefeet,
    finished_living_area = finishedsquarefeet12,
    unfinished_and_finished_area = finishedsquarefeet6,
    total_area = finishedsquarefeet15,
    fireplaces = fireplacecnt,
    garages = garagecarcnt,
    has_spa = hashottuborspa,
    heating_type = heatingorsystemtypeid,
    lot_area = lotsizesquarefeet,
    pools = poolcnt,
    spa_hottub = pooltypeid10,
    pool_with_spa = pooltypeid2,
    pool_no_spa = pooltypeid7,
    county_zoning = propertylandusetypeid,
    city_id = regionidcity,
    county_id = regionidcounty,
    neighborhood_id = regionidneighborhood,
    zipcode_id = regionidzip,
    rooms = roomcnt,
    story_type = storytypeid,
    units = unitcnt,
    patio = yardbuildingsqft17,
    shed = yardbuildingsqft26,
    year_built = yearbuilt,
    stories = numberofstories,
    has_fireplace = fireplaceflag,
    tax_parcel = taxvaluedollarcnt,
    tax_year = assessmentyear,
    tax_land = landtaxvaluedollarcnt,
    tax_total = taxamount,
    tax_pastdue = taxdelinquencyflag,
    tax_pastdue_year = taxdelinquencyyear
  )

# glimpse (df_2016props) # looks better!
```

There are a few columns which have type IDs that are defined in the data dictionary on Kaggle. I'd like to replace the IDs in the current data frame with their values as defined in the data dictionary. Since some of these columns are factor type, I'll have to first convert them back to a character type, update the values, and then re-convert the columns back to factor type.

```{r ID replacement}
df_2016props <- df_2016props %>%
  mutate (AC_type = as.character (AC_type),
          architecture_type = as.character (architecture_type),
          framing_type = as.character (framing_type),
          heating_type = as.character (heating_type),
          county_zoning = as.character (county_zoning),
          story_type = as.character (story_type)
          )

# now that we have character types, replace the IDs with description strings
df_2016props["AC_type"][df_2016props["AC_type"] == "1"] <- "Central"
df_2016props["AC_type"][df_2016props["AC_type"] == "2"] <- "Chilled Water"
df_2016props["AC_type"][df_2016props["AC_type"] == "3"] <- "Evaporative Cooler"
df_2016props["AC_type"][df_2016props["AC_type"] == "4"] <- "Geothermal"
df_2016props["AC_type"][df_2016props["AC_type"] == "5"] <- "None"
df_2016props["AC_type"][df_2016props["AC_type"] == "6"] <- "Other"
df_2016props["AC_type"][df_2016props["AC_type"] == "7"] <- "Packaged AC Unit"
df_2016props["AC_type"][df_2016props["AC_type"] == "8"] <- "Partial"
df_2016props["AC_type"][df_2016props["AC_type"] == "9"] <- "Refrigeration"
df_2016props["AC_type"][df_2016props["AC_type"] == "10"] <- "Ventilation"
df_2016props["AC_type"][df_2016props["AC_type"] == "11"] <- "Wall Unit"
df_2016props["AC_type"][df_2016props["AC_type"] == "12"] <- "Window Unit"
df_2016props["AC_type"][df_2016props["AC_type"] == "13"] <- "Yes"

df_2016props["architecture_type"][df_2016props["architecture_type"] == "1"] <- "A-Frame"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "2"] <- "Bungalow"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "3"] <- "Cape Cod"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "4"] <- "Cottage"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "5"] <- "Colonial"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "6"] <- "Custom"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "7"] <- "Contemporary"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "8"] <- "Conventional"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "9"] <- "Dome"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "10"] <- "French Provincial"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "11"] <- "Georgian"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "12"] <- "High Rise"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "13"] <- "Historical"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "14"] <- "Log Cabin / Rustic"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "15"] <- "Mediterranean"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "16"] <- "Modern"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "17"] <- "Mansion"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "18"] <- "English"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "19"] <- "Other"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "20"] <- "Prefab"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "21"] <- "Ranch / Rambler"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "22"] <- "Raised Ranch"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "23"] <- "Spanish"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "24"] <- "Traditional"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "25"] <- "Tudor"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "26"] <- "Unfinished"
df_2016props["architecture_type"][df_2016props["architecture_type"] == "27"] <- "Victorian"

df_2016props["framing_type"][df_2016props["framing_type"] == "1"] <- "Fireproofed structural steel"
df_2016props["framing_type"][df_2016props["framing_type"] == "2"] <- "Fireproofed concrete"
df_2016props["framing_type"][df_2016props["framing_type"] == "3"] <- "Non-combustible exterior"
df_2016props["framing_type"][df_2016props["framing_type"] == "4"] <- "Wood / wood and steel"
df_2016props["framing_type"][df_2016props["framing_type"] == "5"] <- "Specialized / Other"

df_2016props["heating_type"][df_2016props["heating_type"] == "1"] <- "Baseboard"
df_2016props["heating_type"][df_2016props["heating_type"] == "2"] <- "Central"
df_2016props["heating_type"][df_2016props["heating_type"] == "3"] <- "Coal"
df_2016props["heating_type"][df_2016props["heating_type"] == "4"] <- "Convection"
df_2016props["heating_type"][df_2016props["heating_type"] == "5"] <- "Electric"
df_2016props["heating_type"][df_2016props["heating_type"] == "6"] <- "Forced Air"
df_2016props["heating_type"][df_2016props["heating_type"] == "7"] <- "Floor / Wall"
df_2016props["heating_type"][df_2016props["heating_type"] == "8"] <- "Gas"
df_2016props["heating_type"][df_2016props["heating_type"] == "9"] <- "Geothermal"
df_2016props["heating_type"][df_2016props["heating_type"] == "10"] <- "Gravity"
df_2016props["heating_type"][df_2016props["heating_type"] == "11"] <- "Heat Pump"
df_2016props["heating_type"][df_2016props["heating_type"] == "12"] <- "Hot Water"
df_2016props["heating_type"][df_2016props["heating_type"] == "13"] <- "None"
df_2016props["heating_type"][df_2016props["heating_type"] == "14"] <- "Other"
df_2016props["heating_type"][df_2016props["heating_type"] == "15"] <- "Oil"
df_2016props["heating_type"][df_2016props["heating_type"] == "16"] <- "Partial"
df_2016props["heating_type"][df_2016props["heating_type"] == "17"] <- "Propane"
df_2016props["heating_type"][df_2016props["heating_type"] == "18"] <- "Radiant"
df_2016props["heating_type"][df_2016props["heating_type"] == "19"] <- "Steam"
df_2016props["heating_type"][df_2016props["heating_type"] == "20"] <- "Solar"
df_2016props["heating_type"][df_2016props["heating_type"] == "21"] <- "Space / Suspended"
df_2016props["heating_type"][df_2016props["heating_type"] == "22"] <- "Vent"
df_2016props["heating_type"][df_2016props["heating_type"] == "23"] <- "Wood-Burning"
df_2016props["heating_type"][df_2016props["heating_type"] == "24"] <- "Yes"
df_2016props["heating_type"][df_2016props["heating_type"] == "25"] <- "Zone"

# to be continued...
```